Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2015-12-08 Paper/paper.tex
minor changes to intro

% Macros for text notes and comments
\newcommand{\Note}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\lambdans}{\Lambda_\mathrm{NS}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}

@@

\begin{abstract}
We attempt to answer the following questions: 
\begin{enumerate}
 \item Will ignoring the effect of NS deformability during inspiral and NS disruption during merger (in an NSBH binary) change the efficiency of our searches?
 \item Will not modeling NS disruption during NSBH mergers change the systematic and 
 statistical biases in the recovery of binary parameters?
 \item If we use templates which models NS disruption, what information can we recover about
 the $\Lambda_\mathrm{BH}\sim (R/M)^5_\mathrm{NS}$ parameter?
\end{enumerate}
\end{abstract}

@@

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1'' in 
September, 2015. With both detectors a factor of $3-4$ more sensitive than
their first-generation counterparts, we have access to $\approx 50$ times more 
spacial volume of the universe at every instant. Within a 
$3$ month observing run, these detectors would have access to as
many binary mergers as initial-LIGO detectors did in $\sim 6$ years of 
coincident operation. Through gradual updates, both of these instruments will
reach their design sensitivity by $2019$, at which point they will like detect
around $70$ binary mergers a year~\cite{Abadie:2010cf}.

Coalescing binaries containing a neutron star (NS) and a black hole (BH) are
one of the most important sources of gravitational waves (GW) with the aLIGO
and Virgo instruments, which expect to detect $\mathcal{O}(10)$
of such systems every 
year~\cite{Abadie:2010cf}. Conventional astronomical methods have, to date, made 
observations of NSs with masses between $XX-YY M_\odot$ and spin
$|\vec{\chi}\lesssim 0.05$. On the other hand, indirect observations of BHs 
have placed the masses of stellar-collapse BHs between $5-35M_\odot$, with 
spin angular momenta ranging from low to nearly extremal values (i.e., nearly
as fast as possible - see,
e.g.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for examples of 
nearly extremal estimates of BH spins, and see
Refs.~\cite{McClintock:2013vwa,Reynolds:2013qqa} for recent reviews of
astrophysical BH spin measurements). \prayush{[Do we want to emphasize on BH
spins here?]}.

One question that aLIGO is likely to probe is `what is the nature of
matter at nuclear densities', i.e. what is the equation of state (EoS) of 
neutron stars. The EoS determines the compactness of NSs, which
determines how early in the coalescence will the NS be tidally disrupted
by the gravitational field of the BH, resulting in the formation
of an accretion disk and/or ejection of unbound matter. This makes NSBH
mergers are one of the prime targets of the program for joint detection of
GW and electromagnetic signals. 
% 
The difficulty in discerning the matter effects in an NSBH merger GW signal
lies in distinguishing it reliably from a BHBH merger signal. The earlier the
NS disrupts, the more orbits there are over which the effect of NS disruption
is visible. Where exactly does disruption happen is governed both by the 
mass-ratio of the binary, as well as the spin on the BH. Larger BHs (with
$q=m_\mathrm{BH}/m_\mathrm{NS}\gtrsim 7$), as well as BHs with large retrograde
spins ($\chi_\mathrm{BH} \lesssim 0$), tend to swallow the NS whole due to a
rapid merger (i.e. without it disrupting before the inner-most stable circular
orbit (ISCO) is reached). On the other hand, binaries with mass-ratios 
$q=2-5$ with large pro-grade $\chi_\mathrm{BH}$ can disrupt much before 
merger, leaving their imprint in the GW signal emitted close to $1.7-2$~kHz.
%

>>>>

% Macros for text notes and comments
\newcommand{\Note}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\lambdans}{\Lambda_\mathrm{NS}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}

@@

\begin{abstract}
We attempt to answer the following questions: 
\begin{enumerate}
 \item Will ignoring the effect of NS deformability during inspiral and NS 
 disruption during merger (in an NSBH binary) change the efficiency of our 
 searches?
 \item What systematic biases are introduced by neglecting the NS tidal
 disruption close to merger? 
 \item At what SNR do the systematic biases dominate over the statistical 
 errors in the recovery of binary parameters?
 \item By using templates which models NS deformation (over inspiral) and 
 NS disruption (close to merger), what information can we recover about
 the $\Lambda_\mathrm{BH}\sim (R/M)^5_\mathrm{NS}$ parameter?
\end{enumerate}
\end{abstract}

@@

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1''
mid-September 2015, operating at a factor of $3-4$ higher gravitational-wave (GW) 
strain sensitivity than their first-generation counterparts~\cite{Shoemaker2009}.
Coalescing binaries of
compact objects, i.e. Black Holes (BH) and/or Neutron Stars (NS), are the most
likely targets for GW detection with these ground-based instruments. Over a
$3$-month O1, these detectors would have access to as
many binary mergers as initial-LIGO detectors did over approximately $6$ years
of coincident operation. Both of these instruments will be further upgraded in
phases to reach their design sensitivity by $2018-19$, at which point we expect to
observe $\sim 70$ binary mergers a year~\cite{Abadie:2010cf}.

NS-BH binaries are one of the most important sources of gravitational waves (GW)
with these ground-based instruments, and we expect to detect $\mathcal{O}(10)$
of such systems every year~\cite{Abadie:2010cf}. 
%
One question that aLIGO will likely probe is `what is the nature of
matter at nuclear densities', i.e. what is the equation of state (EoS) of 
neutron stars. The EoS determines the compactness of NSs, which determines
their deformation over the course of inspiral (albeit weakly, at $5$PN order in 
phasing~\prayush{CITE PN TIDAL RESULTS}), as well as how early in the coalescence
process will the NS be disrupted by the tidal field of the BH. Once disrupted,
the NS forms an accretion disk around the BH. Most of this matter disk spirals 
into the BH within a few milliseconds, with a small fraction becoming unbounded
and gets ejected~\prayush{CITE BHNS NR}. It is believed that an accompaning
short (lasting less than $2$ seconds) Gamma-ray Burst (sGRB) might be present
under certain physical conditions~\prayush{CITE NSBH-GRB},
making these systems one of the prime targets for joint electromagnetic-GW
astronomy searches~\prayush{CITE EM-GW}.
In such a situtation, the emitted GW signal
is heavily damped post-disruption, and the quasi-normal modes of the resulting
BH are not excited. Such a GW signal is qualitatively very different from a
BH-BH merger, and one of the main goals of this work is to determine the 
signal strength at which this difference will be quantifiabe by aLIGO.
%
However, studies have shown that for mass-ratio $\mbh/\mns\gtrsim 5$ the BH
will need to have fairly large and positively-aligned (with the orbit) spins
in order for its tidal field to be strong enough to disrupt the NS before the
innermost stable circular orbit (ISCO) is reached~\cite{Foucart:2013psa}.

Conventional astronomical 
methods have, to date, made observations of NSs with masses between 
$XX-YY M_\odot$ and spins $|\vec{\chi}_\mathrm{NS}|\lesssim 0.05$. On the other hand, 
indirect observations of stellar-mass BHs place their masses between 
$5-35M_\odot$, with their spin angular momenta $|\vec{\chi}_\mathrm{BH}|$ 
ranging from small to nearly extremal (Kerr) values (see,
e.g.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for examples of 
nearly extremal estimates of BH spins, and see
Refs.~\cite{McClintock:2013vwa,Reynolds:2013qqa} for recent reviews of
astrophysical BH spin measurements).

The difficulty in discerning the matter effects in an NSBH merger GW signal
lies in distinguishing it reliably from a BHBH merger signal. The earlier the
NS disrupts, the more orbits there are over which the effect of NS disruption
is visible. Where exactly does disruption happen is governed both by the 
mass-ratio of the binary, as well as the spin on the BH. Larger BHs (with
$q=m_\mathrm{BH}/m_\mathrm{NS}\gtrsim 7$), as well as BHs with large retrograde
spins ($\chi_\mathrm{BH} \lesssim 0$), tend to swallow the NS whole due to a
rapid merger (i.e. without it disrupting before the inner-most stable circular
orbit (ISCO) is reached). On the other hand, binaries with mass-ratios 
$q=2-5$ with large pro-grade $\chi_\mathrm{BH}$ can disrupt much before 
merger, leaving their imprint in the GW signal emitted close to $1.7-2$~kHz.
%
